---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Inferential Statistics: Hypothesis Testing

::: {style="text-align:justify"}

Hypothesis testing is a statistical approach that helps determine if there is sufficient evidence to reject a null hypothesis in favor of an alternative hypothesis. In R, various built-in functions allow you to perform different hypothesis tests, including t-tests, chi-squared tests, and ANOVA. In this chapter, we explore commonly used hypothesis testing methods in R, providing an introductory look at their applications and functions.

For a more in-depth understanding, refer to additional resources.

-   [Hypothesis testing, type I and type II errors](https://pmc.ncbi.nlm.nih.gov/articles/PMC2996198/): This paper explores the essentials of hypothesis testing in research, covering characteristics and types of hypotheses, statistical principles, type I and type II errors, effect size, alpha and beta, statistical power, and p-values [@banerjee2009].

-   [Hypothesis Tests](https://pmc.ncbi.nlm.nih.gov/articles/PMC7807926/pdf/main.pdf): This article aims to provide readers with a clear understanding of the purpose of hypothesis testing, guidance on selecting the appropriate test for various research scenarios using a reference table, and insights into interpreting p-values effectively [@walker2019].

:::

## One-Sample T-Test

::: {style="text-align:justify"}

The one-sample t-test is a statistical method that compares the mean of a sample to a known reference value—usually a population mean—to assess if there is a significant difference. In this context, using the NHANES dataset, we can investigate whether the mean blood pressure (BP) of a sample of adults differs significantly from the standard "normal" BP value. For adults, normal BP is often considered to be around 120 mmHg for systolic pressure.

Here’s how we can set up and conduct this test:

**Hypotheses:**

**Null Hypothesis ($H_0$)**: The mean systolic blood pressure of the sample is equal to the normal value (120 mmHg).

**Alternative Hypothesis ($H_1$)**: The mean systolic blood pressure of the sample differs from 120 mmHg.

**R Implementation**: We’ll use the `t.test()` function to conduct the one-sample t-test, specifying 120 as the value for comparison.

```{r}
pacman::p_load(tidyverse)

# Load NHANES data


df <- NHANES::NHANES

df <- df |> 
  janitor::clean_names()

df <- df |> 
  mutate(bp_sys_post = case_when(
    bmi > 25 ~ round(bp_sys_ave + runif(n(), -8, -1), 0),  # Generate a new random number for each row
    TRUE ~ round(bp_sys_ave + runif(n(), -2, 2), 0)        # Same for those with bmi <= 25
  ))



# Filter to include only adults
df <- df |> 
  filter(age >= 18)

# Conduct a one-sample t-test
t_test_result <- t.test(df$bp_sys_ave, mu = 120, na.rm = TRUE)

# View the result
t_test_result


```


**Interpretation of Results:**

- t-Statistic: 3.618

- Degrees of Freedom: 7,204

- p-value: 0.0002981

- 95% Confidence Interval for Mean BP: (120.33, 121.12)

- Sample Mean: 120.73 mmHg

With a p-value of 0.0003 (below 0.05), we conclude that the sample’s average blood pressure of 120.73 mmHg is statistically different from the normal BP value of 120 mmHg. However, for clinicians, it’s essential to think about what this small difference actually means for patient care.

In this case, while there’s a statistically significant difference, the 0.73 mmHg increase from the standard value is likely too small to have clinical importance. This reminds us that even if a result is statistically significant, it’s the practical impact on patient outcomes that ultimately matters.

:::

## Two-sample t-test

::: {style="text-align:justify"}

The **two-sample t-test** is a statistical test used to compare the means of two independent groups. In this context, we want to test whether there is a significant difference in height between males and females using the NHANES dataset.

**Hypotheses:**

**Null Hypothesis ($H_0$)**: There is no difference in the mean height between males and females.

**Alternative Hypothesis ($H_1$)**: : There is a difference in the mean height between males and females.

**Conduct the Test in R**: Assuming we have loaded and cleaned the NHANES dataset, we can perform the two-sample t-test on height.


```{r}
# Conduct two-sample t-test for height difference between males and females
t_test_height <- t.test(height ~ gender, data = df, var.equal = TRUE)

# View the results
t_test_height

```

In the t.test() function in R, the argument var.equal = TRUE specifies that we assume the two groups have equal variances. By default, the two-sample t-test in R does not assume equal variances (this is called Welch's t-test). When var.equal = TRUE, we perform a pooled two-sample t-test, where the variances from each group are combined (or "pooled") to estimate a common variance.

**Interpretation of Results**

- t-Statistic: -80.661

- Degrees of Freedom (df): 7,422

- p-value: < 2.2e-16

- 95% Confidence Interval for the Difference in 
- Means: (-14.15, -13.48)

- Sample Means:

Female: 162.06 cm

Male: 175.87 cm

Conclusion: The very low p-value (< 0.05) indicates that we reject the null hypothesis, suggesting that there is a statistically significant difference in average height between males and females in this sample. The 95% confidence interval suggests that males are, on average, between 13.48 and 14.15 cm taller than females.

:::

## Paired t-test

::: {style="text-align:justify"}

To examine the effect of an intervention, a paired t-test can be used to compare blood pressure (BP) measurements taken pre- and post-intervention among individuals with a BMI greater than 25. The paired t-test is ideal for dependent samples, where each individual has measurements in both conditions (pre and post), allowing us to assess whether the intervention significantly impacted BP.

For the paired t-test examining blood pressure changes before and after an intervention among individuals with a BMI over 25, the hypotheses can be outlined as follows:

**Hypotheses:**

**Null Hypothesis ($H_0$)**: The mean systolic blood pressure before the intervention is equal to the mean blood pressure after the intervention for individuals with a BMI greater than 25. This implies that the intervention had no effect on blood pressure. 

**Alternative Hypothesis ($H_1$)**:  The mean systolic blood pressure before the intervention is different from the mean blood pressure after the intervention for individuals with a BMI greater than 25, suggesting a potential effect of the intervention on blood pressure.

If the test results in a p-value less than our significance threshold (typically 0.05), we would reject the null hypothesis and conclude that there is a statistically significant difference in blood pressure, likely attributable to the intervention.

```{r}

# Filter the dataset for individuals with BMI > 25
df_filtered <- df %>%
  filter(bmi > 25)

# Run a paired t-test
t_test_result <- t.test(
  df_filtered$bp_sys_post,
  df_filtered$bp_sys_ave,
  paired = TRUE)

# Display the result
t_test_result


```


**Paired t-Test Results Interpretation**


The t-statistic is -155.16, indicating a significant difference in systolic blood pressure (BP) between the average pre-intervention and post-intervention measurements.


The degrees of freedom is 4855, indicating a large sample size, which adds robustness to the statistical findings.

The p-value is < 2.2e-16, which is extremely low and suggests a statistically significant difference in systolic blood pressure before and after the intervention. Since this p-value is well below the standard threshold of 0.05, we reject the null hypothesis, indicating that there is a meaningful change in BP levels.
Alternative Hypothesis:

The alternative hypothesis states that the true mean difference in systolic blood pressure between the two measurements (pre- and post-intervention) is not equal to zero.


The 95% confidence interval for the mean difference is (-4.603987, -4.489094). This interval indicates that we can be 95% confident that the true mean decrease in systolic blood pressure lies between -4.60 mmHg and -4.49 mmHg. Since the entire interval is negative, this strongly supports the conclusion that the intervention has led to a significant reduction in systolic BP.


The average difference in systolic blood pressure is approximately -4.55 mmHg, suggesting that, on average, individuals with a BMI greater than 25 experienced a decrease of 4.55 mmHg in systolic BP following the intervention.


The results of the paired t-test indicate that the intervention was effective, resulting in a statistically significant decrease in systolic blood pressure among individuals with a BMI greater than 25. The average decrease of 4.55 mmHg is both statistically significant and clinically relevant. This finding underscores the importance of the intervention in managing blood pressure among this population, suggesting that similar strategies may be beneficial for further lowering systolic blood pressure and improving cardiovascular health.

:::

## Chi-Squared Test

::: {style="text-align:justify"}

The chi-squared test is a statistical method used to determine if there is a significant association between two categorical variables. By comparing the observed frequency distribution of categories with the expected distribution (assuming no association), the test evaluates whether the variables are independent of each other or if there is an association. This test is particularly useful in analyzing relationships in large datasets with categorical data, like survey responses or patient characteristics.

As an example, we’ll use the chi-squared test to examine whether there is an association between BMI category and diabetes status among adults in our dataset.

**Hypotheses:**

**Null Hypothesis ($H_0$)**: There is no association between BMI category and diabetes status (BMI and diabetes status are independent).

**Alternative Hypothesis ($H_1$)**:  There is an association between BMI category and diabetes status.

**Data Preparation**

Ensure that the bmi_who and diabetes variables are factors:

If not then make it factor variable.

```{r}
class(df$bmi_who)

class(df$diabetes)
```

Both variables are factor.


**Creating a Contingency Table in Tidy Format**

To perform a chi-squared test, we first need a contingency table that shows the counts of individuals in each combination of bmi_who and diabetes categories. 

```{r}
# Create contingency table in tidy format
contingency_table <- table(df$bmi_who, df$diabetes)

```

**Performing the Chi-Squared Test**

To assess whether an association exists between BMI category and diabetes, we apply the chi-squared test as follows:

```{r}
# Conduct the chi-squared test
chi_test <- chisq.test(contingency_table)

# View the test results
chi_test


```


**Interpretation of Chi-Squared Test Results**

- Chi-Squared Statistic (X-squared): 206.12

- Degrees of Freedom (df): 3

- p-value: < 2.2e-16

Since the p-value is extremely small (less than 0.05), we reject the null hypothesis, indicating that there is a statistically significant association between BMI category and diabetes status. This suggests that diabetes prevalence differs across BMI categories in our sample. 

However, while statistical significance tells us there is an association, it does not reveal the direction or strength of the relationship. To determine the strength and direction—such as whether higher BMI categories are positively associated with increased diabetes risk—you could perform a regression analysis.

For binary outcomes like diabetes status, logistic regression would be a suitable approach. It would allow us to model the probability of diabetes as a function of BMI categories while providing insights into how each category impacts the likelihood of diabetes, as reflected by odds ratios. This approach also enables adjustments for potential confounders, offering a more comprehensive understanding of the relationship.

:::

## One-way ANOVA 

::: {style="text-align:justify"}

One-way ANOVA (Analysis of Variance) is used to compare the means of more than two groups to determine if at least one group mean differs significantly from the others. This method is suitable when we have a categorical independent variable (factor) with multiple levels and a continuous dependent variable. For example, we could use one-way ANOVA to test if the average blood pressure differs across BMI categories.

As an example, we’ll use the one-way ANOVA To determine if average systolic blood pressure varies across BMI categories.

**Hypotheses:**

**Null Hypothesis ($H_0$)**: The average systolic blood pressure is the same across all BMI categories.

**Alternative Hypothesis ($H_1$)**:  The average systolic blood pressure differs for at least one BMI category.

```{r}
# Fit the ANOVA model
anova_model <- aov(bp_sys_ave ~ bmi_who, data = df)

# View the summary
summary(anova_model)

```

**Interpretation of One-Way ANOVA Results**

- **Degrees of Freedom (Df)**: The degrees of freedom for the BMI categories is 3, meaning there are four BMI groups. The residual degrees of freedom is 7116, accounting for the total sample size minus the number of groups.

- **Sum of Squares (Sum Sq)**: This indicates the variability within each group (Residuals) and between the groups (BMI categories). A larger Sum of Squares for the BMI groups relative to the residuals suggests substantial variation in systolic blood pressure due to BMI category.

- **Mean Squares (Mean Sq)**: This is the Sum of Squares divided by the respective degrees of freedom. The larger Mean Square for BMI groups indicates a larger variability attributed to BMI categories compared to random error.

- **F-value**: The F-value of 50.35 is substantial, suggesting that the variation in systolic blood pressure between the BMI categories is much greater than within each category.

- **p-value (< 2e-16)**: The extremely small p-value is well below 0.05, leading us to reject the null hypothesis. This indicates that there is a statistically significant difference in systolic blood pressure among the different BMI categories in this sample.

Since the ANOVA indicates a significant difference, a post-hoc analysis (such as Tukey's HSD) could be conducted to identify which BMI categories specifically differ from each other in terms of average systolic blood pressure.

:::


## References {.unnumbered}